Unshuffling Permutations

### 2015-09-22 submission.tex
add figure

% \renewcommand{\floatpagefraction}{.9}
% \renewcommand{\textfraction}{.1}
\renewcommand{\arraystretch}{1.2}

@@

@@

%%
%%
%% ---- Introduction ----
%%
\section{Introduction}
\label{section:Introduction}
The {\em shuffle product} $\shuffle$ is a very famous operation on words
first defined by Eilenberg and Mac Lane~\cite{Eilenberg:MacLane:1953}.
Given three words $u$, $v_1$, and $v_2$, $u$ is said to be a \emph{shuffle}
of $v_1$ and $v_2$ if it can be formed by interleaving the letters from
$v_1$ and $v_2$ in a way that maintains the left-to-right ordering of the
letters from each word. In addition to purely combinatorial questions,
the shuffle product of words leads to the following computational ones:
\begin{enumerate}
    \item \label{item:problem_1}
    Given two words $v_1$ and $v_2$, compute the set $v_1 \shuffle v_2$.
    \item \label{item:problem_2}
    Given three words $u$, $v_1$, and $v_2$, decide if $u$ is a shuffle
    of $v_1$ and $v_2$.
    \item \label{item:problem_3}
    Given words $u$, $v_1$, \dots, $v_k$, decide if $u$ is in
    $v_1 \shuffle \dots \shuffle v_k$.
    \item \label{item:problem_4}
    Given a word $u$, decide if there is a word $v$ such that $u$ is
    in $v \shuffle v$.
\end{enumerate}
Even if these problems seem similar, they are very different in terms
of complexity. Let us now review some facts about these. In what follows,
$n$ denotes the size of $u$ and $m_i$ denotes the size of each $v_i$.
A solution to Problem~\ref{item:problem_1} can be computed in
\begin{equation}
    O\left((m_1 + m_2) \; \binom{m_1 + m_2}{m_1}\right)
\end{equation}
time~\cite{Spehner:TCS:1986}. An improvement and a generalization
of~Problem~\ref{item:problem_1} has been proposed
in~\cite{Allauzen:IGM:2000}, where it is proved that given words
$v_1$, \dots, $v_k$, the iterated shuffle
$v_1 \shuffle \dots \shuffle v_k$ can be computed in
\begin{equation}
    O\left(\binom{m_1 + \dots + m_k}{m_1, \dots, m_k}\right)
\end{equation}
time. Problem~\ref{item:problem_2} is in \Pclass; it is indeed a
classical textbook exercise to design an efficient dynamic programming
algorithm solving it. It can be tested in $O\left(n^2 / \log(n)\right)$
time~\cite{Leeuwen:Nivat:IPL:1982}. To the best of our knowledge, the
first $O(n^2)$ time algorithm for this problem appeared
in~\cite{Mansfield:DAM:1983}. This algorithm can easily be extended to
check in polynomial-time whether or not a word is in the shuffle of any
fixed number of given words. Nevertheless, Problem~\ref{item:problem_3}
is \NPC~\cite{Mansfield:DAM:1983,Warmuth:Haussler:JCSS:1984}. This
remains true even if the ground alphabet has size
$3$~\cite{Warmuth:Haussler:JCSS:1984}. Of particular interest,
it is shown in~\cite{Warmuth:Haussler:JCSS:1984} that
Problem~\ref{item:problem_3} remains \NPC even if all the words $v_i$,
$i \in [k]$, are identical, thereby proving that, for two words $u$ and
$v$, it is \NPC to decide whether or not $u$ is in the iterated shuffle
of $v$. Again, this remains true even if the ground alphabet has size $3$.
Let us now finally focus on Problem~\ref{item:problem_4}. It is shown
in~\cite{Buss:Soltys:2014,Rizzi:Vialette:CSR:2013} that it is \NPC to
decide if a word $u$ is a \emph{square} (w.r.t. the shuffle), that is
a word $u$ with the property that there exists a word $v$ such that $u$
is a shuffle of $v$ with itself. Hence, Problem~\ref{item:problem_4}
is \NPC.
\smallskip

@@

Let us now provide some definitions about graphs and matching, used in
the sequel. For a graph $G$, we denote $\VS{G}$ as the set of vertices
and $\ES{G}$ as the set of edges of $G$. Let $u = u_1 u_2 \dots u_n$ be
a word. The \emph{clique graph associated to $u$}, denoted by $K_u$, is
the complete graph on $\VS{K_u} = \{u_1, u_2, \ldots, u_n\}$. The
structure of this underlying graph is linear, \emph{i.e.}, the set of
its vertices is equipped with a natural total order $<$ defined by
$u_i < u_j$ if and only if $i < j$. For a linear graph, we write
$(u_i, u_j)$ instead of $\{u_i, u_j\}$ for an edge if it is clear from
the context that $i < j$. Recall that two edges of a graph are
\emph{independent} if they do not share a common vertex, and that a
\emph{matching} $\mathcal{M}$ in $G$ is a set of pairwise independent
edges. A matching is \emph{perfect} if it covers all the vertices of the
graph. In case the set of vertices is equipped with a total order, a
matching $\mathcal{M}$ is said to be \emph{containment-free} if there do
not exist (independent) edges $(u_i, u_j)$ and $(u_k, u_\ell)$ in
$\mathcal{M}$ such that $u_i < u_k < u_\ell < u_j$ or
$u_k < u_i < u_j < u_\ell$.

@@

The purpose of this section is to define a shuffle product $\SHUFFLE$
on permutations. To define this shuffle product, we shall first define
a coproduct denoted by $\Delta$, enabling to unshuffle permutations.
By duality, $\Delta$ implies the definition of $\SHUFFLE$. The reason
why we need to pass by the definition of $\Delta$ to define $\SHUFFLE$
is justified by the fact that a lot of properties of $\SHUFFLE$ depend
of properties of $\Delta$ and that this strategy allows to write concise
and clear proofs of them. We invite the reader unfamiliar with the
concepts of coproduct and duality to consult~\cite{Joni:Rota:1979}
or~\cite{Grinberg:Reiner:2014}.

@@

    \begin{claim} \label{claim:square_permutation_to_square_binary}
        If $\pi$ is a square permutation avoiding $213$ and $231$,
        $\PERMTOBIN(\pi)$ is a square binary word.
    \end{claim}
\end{proof}

@@

This section is devoted to proving hardness of recognizing square
permutations. In the same way as happens with words, we shall use a
linear graph framework where deciding whether a permutation is a square
reduces to computing some specific matching in the associated linear
graph~\cite{Buss:Soltys:2014,Rizzi:Vialette:CSR:2013}. It reads as follows.

\begin{lemma}
  \label{lemma:matching}
  Let $\pi$ be a permutation. The following statements are equivalent:
  \begin{enumerate}
    \item The permutation $\pi$ is a square.
    \item There exists a perfect containment-free matching $\mathcal{M}$
    of $G_\pi$ such that for any two distinct edges $(a, b)$ and
    $(a', b')$ in $\mathcal{M}$ we have $\pi(a) < \pi(a')$ if and only
    if $\pi(b) < \pi(b')$.
  \end{enumerate}
\end{lemma}

>>>>

% \renewcommand{\floatpagefraction}{.9}
% \renewcommand{\textfraction}{.1}
\renewcommand{\arraystretch}{1.2}
\newcommand{\rt}{\textcolor{red}}

@@

% 1 2 2 1
\newcommand{\MWithContainment}{%
\begin{tikzpicture}[scale=0.1,inner sep=.8pt,node distance=.25cm]
\draw node [draw,circle,fill=black] (U1)               {};
\draw node [draw,circle,fill=black] [right of=U1] (U2) {};
\draw node [draw,circle,fill=black] [right of=U2] (U3) {};
\draw node [draw,circle,fill=black] [right of=U3] (U4) {};
\draw (U2.north) .. controls ($ (U2.north) + (0,1.75) $) and ($ (U3.north) + (0,1.75) $) .. (U3.north);
\draw (U1.north) .. controls ($ (U1.north) + (0,2.5) $) and ($ (U4.north) + (0,2.5) $) .. (U4.north);
\end{tikzpicture}
}% end \MWithContainment

@@

%%
%%
%% ---- Introduction ----
%%
\section{Introduction}
\label{section:Introduction}
The {\em shuffle product}, denoted $\shuffle$, is a very famous operation on words
first defined by Eilenberg and Mac Lane~\cite{Eilenberg:MacLane:1953}.
Given three words $u$, $v_1$, and $v_2$, $u$ is said to be a \emph{shuffle}
of $v_1$ and $v_2$ if it can be formed by interleaving the letters from
$v_1$ and $v_2$ in a way that maintains the left-to-right ordering of the
letters from each word.
Besides purely combinatorial questions,
the shuffle product of words leads to the following computational ones:
\begin{enumerate}
    \item \label{item:problem_1}
    Given two words $v_1$ and $v_2$, compute the set $v_1 \shuffle v_2$.
    \item \label{item:problem_2}
    Given three words $u$, $v_1$, and $v_2$, decide if $u$ is a shuffle
    of $v_1$ and $v_2$.
    \item \label{item:problem_3}
    Given words $u$, $v_1$, \dots, $v_k$, decide if $u$ is in
    $v_1 \shuffle \dots \shuffle v_k$.
    \item \label{item:problem_4}
    Given a word $u$, decide if there is a word $v$ such that $u$ is
    in $v \shuffle v$.
\end{enumerate}
Even if these problems seem similar, they are very different in terms
of complexity. Let us now review some facts about these. In what follows,
$n$ denotes the size of $u$ and $m_i$ denotes the size of each $v_i$.
A solution to Problem~\ref{item:problem_1} can be computed in
\begin{equation}
    O\left((m_1 + m_2) \; \binom{m_1 + m_2}{m_1}\right)
\end{equation}
time~\cite{Spehner:TCS:1986}. An improvement and a generalization
of~Problem~\ref{item:problem_1} has been proposed
in~\cite{Allauzen:IGM:2000}, where it is proved that given words
$v_1$, \dots, $v_k$, the iterated shuffle
$v_1 \shuffle \dots \shuffle v_k$ can be computed in
\begin{equation}
    O\left(\binom{m_1 + \dots + m_k}{m_1, \dots, m_k}\right)
\end{equation}
time. Problem~\ref{item:problem_2} is in \Pclass; it is indeed a
classical textbook exercise to design an efficient dynamic programming
algorithm solving it. It can be tested in $O\left(n^2 / \log(n)\right)$
time~\cite{Leeuwen:Nivat:IPL:1982}. To the best of our knowledge, the
first $O(n^2)$ time algorithm for this problem appeared
in~\cite{Mansfield:DAM:1983}. This algorithm can easily be extended to
check in polynomial-time whether or not a word is in the shuffle of any
fixed number of given words. Nevertheless, Problem~\ref{item:problem_3}
is \NPC~\cite{Mansfield:DAM:1983,Warmuth:Haussler:JCSS:1984}. This
remains true even if the ground alphabet has size
$3$~\cite{Warmuth:Haussler:JCSS:1984}. Of particular interest,
it is shown in~\cite{Warmuth:Haussler:JCSS:1984} that
Problem~\ref{item:problem_3} remains \NPC even if all the words $v_i$,
$i \in [k]$, are identical, thereby proving that, for two words $u$ and
$v$, it is \NPC to decide whether or not $u$ is in the iterated shuffle
of $v$. Again, this remains true even if the ground alphabet has size $3$.
Let us now finally focus on Problem~\ref{item:problem_4}. It is shown
in~\cite{Buss:Soltys:2014,Rizzi:Vialette:CSR:2013} that it is \NPC to
decide if a word $u$ is a \emph{square} (w.r.t. the shuffle), that is
a word $u$ with the property that there exists a word $v$ such that $u$
is a shuffle of $v$ with itself. Hence, Problem~\ref{item:problem_4}
is \NPC.
\smallskip

@@

Let us now provide some definitions about graphs and matching that are used in
the sequel. For a graph $G$, we denote $\VS{G}$ as the set of vertices
and $\ES{G}$ as the set of edges of $G$. Let $u = u_1 u_2 \dots u_n$ be
a word. The \emph{clique graph associated to $u$}, denoted by $K_u$, is
the complete graph on $\VS{K_u} = \{u_1, u_2, \ldots, u_n\}$. The
structure of this underlying graph is linear, \emph{i.e.}, the set of
its vertices is equipped with a natural total order $<$ defined by
$u_i < u_j$ if and only if $i < j$. For a linear graph, we write
$(u_i, u_j)$ instead of $\{u_i, u_j\}$ for an edge if it is clear from
the context that $i < j$. Recall that two edges of a graph are
\emph{independent} if they do not share a common vertex, and that a
\emph{matching} $\mathcal{M}$ in $G$ is a set of pairwise independent
edges. A matching is \emph{perfect} if it covers all the vertices of the
graph.
For an edge $e = \{u,v\}$, we shall occasionally say that $u$ and $v$
are \emph{involved} in $e$.
In case the set of vertices is equipped with a total order, a
matching $\mathcal{M}$ is said to be \emph{containment-free} if there do
not exist (independent) edges $(u_i, u_j)$ and $(u_k, u_\ell)$ in
$\mathcal{M}$ such that $u_i < u_k < u_\ell < u_j$ or
$u_k < u_i < u_j < u_\ell$
(\emph{i.e.}, $\mathcal{M}$ avoids \MWithContainment).

@@

The purpose of this section is to define a shuffle product $\SHUFFLE$
on permutations. To define this shuffle product, we shall first define
a coproduct denoted by $\Delta$, enabling to unshuffle permutations.
By duality, $\Delta$ implies the definition of $\SHUFFLE$. The reason
why we need to pass by the definition of $\Delta$ to define $\SHUFFLE$
is justified by the fact that a lot of properties of $\SHUFFLE$ depend
of properties of $\Delta$, and that this strategy allows to write concise
and clear proofs of them. We invite the reader unfamiliar with the
concepts of coproduct and duality to consult~\cite{Joni:Rota:1979}
or~\cite{Grinberg:Reiner:2014}.

@@

    \begin{claim} \label{claim:square_permutation_to_square_binary}
        If $\pi$ is a square permutation avoiding $213$ and $231$,
        $\PERMTOBIN(\pi)$ is a square binary word.
    \end{claim}
    \qed
\end{proof}

@@

This section is devoted to proving hardness of recognizing square
permutations. In the same way as happens with words, we shall use a
linear graph framework where deciding whether a permutation is a square
reduces to computing some specific matching in the associated linear
graph~\cite{Buss:Soltys:2014,Rizzi:Vialette:CSR:2013}. It reads as follows
(see Fig.\ref{fig:example containment-free matching}).

\begin{lemma}
  \label{lemma:matching}
  Let $\pi$ be a permutation. The following statements are equivalent:
  \begin{enumerate}
    \item The permutation $\pi$ is a square.
    \item There exists a perfect containment-free matching $\mathcal{M}$
    of $K_\pi$ such that for any two distinct edges $(a, b)$ and
    $(a', b')$ in $\mathcal{M}$ we have $\pi(a) < \pi(a')$ if and only
    if $\pi(b) < \pi(b')$.
  \end{enumerate}
\end{lemma}

\begin{figure}[ht!]
  \centering
  \begin{tikzpicture}[scale=0.5,inner sep=2pt,node distance=1cm]
    \draw node (K) {$K_\pi$};
    \draw node [draw, circle, minimum size=.55cm, fill=black!10] [right of=K]   (U01) {$2$};
    \draw node [draw, circle, minimum size=.55cm, fill=black!10] [right of=U01] (U02) {$A$};
    \draw node [draw, circle, minimum size=.55cm] [right of=U02] (U03) {$4$};
    \draw node [draw, circle, minimum size=.55cm, fill=black!10] [right of=U03] (U04) {$7$};
    \draw node [draw, circle, minimum size=.55cm, fill=black!10] [right of=U04] (U05) {$1$};
    \draw node [draw, circle, minimum size=.55cm] [right of=U05] (U06) {$9$};
    \draw node [draw, circle, minimum size=.55cm] [right of=U06] (U07) {$5$};
    \draw node [draw, circle, minimum size=.55cm] [right of=U07] (U08) {$3$};
    \draw node [draw, circle, minimum size=.55cm, fill=black!10] [right of=U08] (U09) {$8$};
    \draw node [draw, circle, minimum size=.55cm] [right of=U09] (U10) {$6$};
    \draw [thick]
      (U01.north) .. controls ($ (U01.north) + (0,2.25) $) and ($ (U03.north) + (0,2.25) $) .. (U03.north);
    \draw [thick]
      (U02.north) .. controls ($ (U02.north) + (0,2.25) $) and ($ (U06.north) + (0,2.25) $) .. (U06.north);
    \draw [thick]
      (U04.north) .. controls ($ (U04.north) + (0,2.5) $) and ($ (U07.north) + (0,2.5) $) .. (U07.north);
    \draw [thick]
      (U05.north) .. controls ($ (U05.north) + (0,2.5) $) and ($ (U08.north) + (0,2.5) $) .. (U08.north);
    \draw [thick]
      (U09.north) .. controls ($ (U09.north) + (0,2.5) $) and ($ (U10.north) + (0,2.5) $) .. (U10.north);
  \end{tikzpicture}
  \caption{\label{fig:example containment-free matching}%
    The linear graph $K_\pi$ associated to the permutation
    $\pi = 2A47195386$ together with
    a containment-free perfect matching $\mathcal{M}$
    such that for any two distinct edges $(a, b)$ and
    $(a', b')$ in $\mathcal{M}$ we have $\pi(a) < \pi(a')$ if and only
    if $\pi(b) < \pi(b')$.
    From $\mathcal{M}$, it follows that $\pi$ is square as it is
    the shuffle of $2A718$ and $49536$ that are both
    order-isomorphic to $25314$.
  }% end caption
\end{figure}
